Lightweight Real-Time Feature Monitoring

### 2020-07-14 paper.tex
New paper template, new experiment figures

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
Many real-time stream monitoring systems are static once deployed in a production environment. Over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates. Data pattern shift detection refers to the process of finding patterns in data that do not conform to expected or usual behavior. Accurate and timely detection of data pattern deviations allows for immediate measures to be taken. Thus, the problem at hand is to determine when to reconfigure the system, \textit{e.g.}, a Machine Learning model, based on an analysis of the drifts in the stream of data. In this paper, we present a method that makes use of lightweight streaming aggregations and distribution divergence functions to alert about deviations in data patterns in real-time. We generate alerts based on a user-defined probability threshold, for each of the event's fields, also denominated features in our context. We evaluate our method through a series of tests, some with synthetic datasets and some with real data. We further split the tests into single and multi-feature analysis. Our method accurately detected the introduced anomalies in the experiments using synthetic datasets while maintaining high throughput, for single and multi-feature analysis. However, experiments with real data were not as accurate. Despite not knowing the specific reasons for that, we defer further investigation to future work and formulate a set of hypotheses that might explain it and hence are worthy of pursuing next. Our set of experiments supports the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.
\end{abstract}

@@

\section{Method}
\label{sec:Method}
A data stream is a continuous collection of events, each timestamped and with multiple fields. We denote each of the fields as features. Our goal is to detect data pattern deviations for each feature in a streaming fashion. Such shifts will be measured between a static in time reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.
\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/timeperiods.png}
      \caption[]{Reference and streaming timelines}
      \label{fig:timelines}
    \end{center}
\end{figure}
To detect data pattern shifts, we do not need to directly compare the entire window contents of the reference and sliding windows. This approach would require storing all window contents hence making the memory consumption grow linearly with target sliding window and/or reference period size. Instead, we want to use a \textit{(a)} fixed cost in-memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

One aggregation that encodes distributions properly are histograms. Histograms are a representation of the distribution of numerical data thus being ideal to encode the distribution of feature values, satisfying \textit{(c)}. Histograms require storing a fixed-size list of bins and corresponding counts per bin regardless of the volume of data, satisfying property \textit{(a)}. Adding or removing an element to the histogram aggregation is tantamount to determining the correct bin and incrementing or decrementing the associated count, respectively, thus satisfying property \textit{(b)}. Recall Section \ref{sec:back-swag-algs} where we discussed generic sliding window aggregation algorithms and remember the Subtract-On-Evict (SOE) algorithm (Algorithm \ref{pseudo:soe}). Computing a histogram aggregation over a sliding window of size $W$ using SOE implies storing all $W$ elements in memory to know the element to remove after a new one is inserted. This violates our constant in-memory property \textit{(a)} which is why we chose to use the Exponential Moving Averages (EMAs) analyzed in Section \ref{sec:emas}. EMA-like histograms computed over sliding windows do not require the storage of window contents. An EMA-like histogram is \textit{(a)} constant in memory, \textit{(b)} maintained in real-time and \textit{(c)} encodes the distribution of each feature values. These EMA-like histograms are further explained in Section \ref{sec:ema-hist}.

Since the reference window is static, \textit{i.e.}, it is not a sliding window, we can use exact histogram aggregations. Hence, for each feature, we build an exact reference histogram aggregation. On the other hand, in streaming, we use one EMA-like approximate histogram aggregation per feature (Figure \ref{fig:timelines-hists}).

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/timelines-hists.png}
      \caption[]{Reference and Streaming timelines and corresponding histogram aggregations}
      \label{fig:timelines-hists}
    \end{center}
\end{figure}

To compare each feature's reference and target sliding window histogram aggregation we need a measure of divergence. We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein \cite{EncyclopaediaMathematics}. The JSD takes as input two histograms and outputs a value between 0 and 1. The closer this value is to 0 the more similar the distributions are. On the other hand, a JSD value of 1 represents two different distributions. We chose to use the JSD metric due to its simplicity and because it was used in previous work successfully \cite{SAMM}, leaving as future work other possible tests.

Given a divergence or distance value for the two aggregations (reference and target) we need a measure of normal statistical fluctuations. Since we consider the reference period as the normal state we can find the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis. This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand-picked threshold for the divergence value (details in Section \ref{sec:sampling-batch}).

To implement the proposed methodology, we devised a two-phase method. The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section \ref{sec:batch-phase}). The second phase (detailed in Section \ref{sec:stream-phase}) focuses on efficiently updating the target sliding window histogram for each feature and periodically performing a divergence test between the static reference aggregation and the target sliding one, alerting for changes if need be. The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.

\subsection{Exponential Moving Average Histogram} \label{sec:ema-hist}
As stated, we compute histogram aggregations for the reference and target sliding windows. The computation of the reference histogram is done once in batch. Since the reference data is fixed and does not change over time, computing this histogram in batch is more efficient and easier to implement in a distributed system like Spark. In addition, a batch computation also allows us to compute an exact reference histogram. However, the target sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.

In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they compute sliding window aggregations without actually storing the window contents. Definition \ref{def:tuple-ema} shows how to compute an EMA using its recursive formula. An EMA state or value depends only on the previous one multiplied by an exponential decay. Recall that a tuple-based decay is given by:
\[ 2^{- \frac{1}{n_{1/2}}} \]
where $n_{1/2}$ is commonly denoted as a tuple-based half-life.

Algorithm \ref{pseudo:ema-hist} shows how we update our EMA histogram. We assume we have access to global \textit{counts} and \textit{bins} arrays and use a suppression factor of $2^{- \frac{1}{n_{1/2}}}$ with $n_{1/2}$ as the tuple-based half-life. 
The histogram aggregation is updated for each new streaming value \texttt{val} using the \texttt{processValue(val)} function, that increments the correct histogram bin (note that the \textit{bins} array defines the bin boundaries while the \textit{counts} array actually holds each bin's count) and applies the EMA suppression to every bin.
\begin{algorithm}[!htb]
    \caption[EMA histogram]{EMA histogram sliding window emulation}
    \label{pseudo:ema-hist}
    \begin{algorithmic}[1]
        \Function{getBin}{val}
            \State \Return{binarySearch(bins, val)}
        \EndFunction

        \Function{processValue}{val}
            \State $bin \gets getBin(val)$
            \State $counts[bin] \gets counts[bin] + 1$ \Comment{// Increment correct bin}

            \ForEach {$count \in counts$} \Comment{// Apply EMA suppression to all bins}
                \State $count \gets count \times 2^{- \frac{1}{n_{1/2}}}$
            \EndFor
        \EndFunction
    \end{algorithmic}
\end{algorithm}

After processing \textit{x} tuples the oldest tuple is \emph{smoothed} by a factor of $2^{- \frac{x}{n_{1/2}}}$. If we process $x = 4n_{1/2}$ tuples the oldest tuple would have its contribution reduced by
\[ 2^{- \dfrac{4n_{1/2}}{n_{1/2}}} = 2^{-4} \]
which is approximately 0.06. In other words, the \textit{x-th} event would have 6\% contribution towards the EMA state or value meaning older events would have even less weight.

We choose to emulate windows with four times the half-life size (or $4n_{1/2}$), meaning we discard events whose contribution is less than 6\%. We set the half-life to be one-quarter of the size of the window we want to emulate. The size of the window we want to monitor varies from situation to situation, but in the limited scope of this thesis we set to monitor a period extending up to one month (at an error level of about 6\%) so we set the half-life to be the number of tuples that correspond to one week worth of data. This will give more weight to the most recent week but it will include contributions decaying to 6\% in the fourth oldest week.

\subsection{Batch Analysis Phase} \label{sec:batch-phase}
The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and EMA-like aggregations. In other words, we:

\begin{enumerate}
    \item Obtain a reference histogram for each feature using all data in the reference period

    \item Compute the distribution of distances between the reference histogram and random windows of data and corresponding histograms
\end{enumerate}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/ref-hist.png}
      \caption[]{Reference equal-height histogram}
      \label{fig:ref-hist}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.3]{figures/ref-bins.png}
      \caption[]{Reference equal-height histogram}
      \label{fig:ref-hist-bins}
    \end{center}
\end{figure}

Each of the histograms is an EMA-like histogram. An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}. Each sample is processed event by event and a discount factor is applied per event to each bin (EMA expiration factor), thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample. This procedure is illustrated in Figure \ref{fig:EMA-hist-build}.
\begin{figure}[!htb]
    \begin{center}
    %\hspace{2cm}
      \includegraphics[scale=0.8]{figures/ema-hist.png}
      \caption[]{Reference equal-height histogram}
      \label{fig:EMA-hist-build}
    \end{center}
\end{figure}
We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period. We use an approximated histogram to mimic the target approximated histogram we will incrementally maintain in the streaming environment.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/compute-sample-distances.png}
      \caption[Compute sample's distance values]{Compute as many distance values as samples, one for each sample-reference histogram pair}
      \label{fig:compute-sample-distances}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.2]{figures/skewed-left-distro.png}
      \caption[]{Left skewed distribution of distance values}
      \label{fig:skewed-left-distro}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.2]{figures/skewed-right-distro.png}
      \caption[]{Right skewed distribution of distance values}
      \label{fig:skewed-right-distro}
    \end{center}
\end{figure}

@@

\subsection{Streaming Phase} \label{sec:stream-phase}
In the streaming or online phase, the system will process event by event in a true sliding window fashion (recall Section \ref{sec:windows}) and perform a change detection test periodically. This phase makes use of all of the batch phase artifacts, namely, for each feature, the reference histogram, the distribution of distance values and the last sample's histogram.

For each feature, we initialize the approximate histogram with the last sample's data, from the batch phase. Then, for each incoming event and each of the event's features, we update our streaming approximate histogram for that feature by adding the new value and then "cutting a slice" from the top of the histogram bars to simulate eviction, as described in Section \ref{sec:ema-hist}. This is all we do, regarding event processing.

In order to obtain alerts, we periodically perform a multi-feature change detection test. The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the target sliding window histogram of the same feature. This distance value is then tested against the known distribution of distance values for feature \textit{x}. In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile, \textit{e.g.}, the 99th-percentile. When we perform the test, for each feature \textit{x}, we compute the percentile of the distance measured between the reference and target sliding window histogram aggregations. The probability value or \textit{p-value} of a distance value is:

\[\textit{p-value\textsubscript{distance\textsubscript{x}}} = 1 - percentile(\textit{distance\textsubscript{x}})\]

For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01. Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
\[ \textit{percentile(distance\textsubscript{x})} > 1 - \alpha \]
or:
\begin{equation}
    \label{eq:alert-test}
    \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha
\end{equation}

\subsubsection*{Multiple Test Correction} \label{sec:multi-test}
Note that we periodically apply this test (Equation \ref{eq:alert-test}) for each feature, effectively repeating the analysis. When considering several hypotheses, the problem of multiplicity arises: the more hypotheses are checked, the higher the probability of obtaining false positives. In other words, when repeating a test multiple times the multiple test or simultaneous statistical inference problem \cite{MultiTestProblem-Pubmed, MultiTestProblem-Dickhaus2014, MultiTestProblem-Miller1966, multiple-test-correction-geoffrey} arises which states that the more inferences are made, the more likely erroneous inferences are to occur. In the literature, many multiple test correction methods are proposed, but we opted to use the Holm-Bonferroni multiple test correction because it is a simple and easy to compute correction that can reduce the number of false negatives while having as target to control the number of false positives. Furthermore, it does not require to assume any independence between the tested hypothesis (as other methods do) which would almost surely be a wrong assumption for realistic datasets. Therefore we leave as future work the testing of other correction methods. In statistics, Family-Wise Error Rate (FWER) \cite{MultivariateMT, MultitestTamhane2018AdvancesIP} is the probability of having one or more false positives when performing multiple hypotheses tests, \textit{i.e.}, considering one hypothesis true when it is not. The Holm-Bonferroni method \cite{HolmBonferroni} is one of many approaches for controlling the Family-Wise Error Rate \textit{(FWER)} by adjusting the rejection criteria for each hypothesis.

\subsubsection*{Holm-Bonferroni Correction} \label{sec:holmbonferroni}
The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature. This correction is done by dividing the threshold $\alpha$ by a correction factor, as described next.

First, we compute the \textit{p-value} of each feature. We then order our features by \textit{p-value}, in ascending order. Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if:
\begin{equation}
    p-value_{distance_{x}} < \alpha / (m - k)
    \label{eq:corrected-pvalue}
\end{equation}
with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature. We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and alerts are raised.

\subsection{Final Remarks}

\subsubsection*{Minimum Number of Samples} \label{sec:nsamples}
In Section \ref{sec:sampling-batch} we stated we made \textit{S} samples in the batch phase. But what is the optimal number of samples to make? How can we mathematically derive a formula for the number of samples to make and avoid \textit{ad hoc} user-defined constants?

When checking for alerts we are performing one test for each feature, hence we perform \textit{n\textsubscript{features}} tests. We apply a multiple test correction on the tests performed and want to set the FWER to $\alpha$. In other words, we want to alert deviations for the full set of tests globally that have probability $\alpha$ < 1\%, for example.

We use the Holm-Bonferroni correction and, for each feature, we need to reliably estimate the upper tail of the distribution which corresponds to the \textit{$\alpha$ / n\textsubscript{features}} region. Each sample we do translates into a distance value and the collection of these data points builds the distribution of distance values. We want to be sure that the number of samples \textit{S} that we do generates enough distance points to have a high probability of having one or more points in the upper tail region. To that end, we define the probability of not having at least one point in the upper tail of the distance distribution as $\gamma$.

In \cite{SAMM}, the authors prove that the minimum number of samples $S$ needed to satisfy these probability requirements is:
\begin{equation}
    \label{eq:optimal-n-samples}
    \textit{S} = \ceil[\bigg]{\frac{\log\gamma}{log(1 - \frac{\alpha}{n_{features}})}}
\end{equation}

\subsubsection*{Method classification}
Similarly to the methods presented in Sections \ref{kifer} and \ref{sec:SAMM}, we use a two-windowed model that can take uni or multivariate input. However, we perform a multivariate analysis of all feature's time-series with the use of the Holm-Bonferroni correction. Hence, we say our method is multivariate. We also alert entire windows or subsequences as anomalous. Thus, according to the taxonomy presented in Section \ref{sec:outliers}, our method is a dissimilarity-based multivariate subsequence outlier detection method.

\subsubsection*{Time complexity of the streaming phase}
The critical phase of our method is the stream analysis because it has to process new events and alert for possible deviations in real-time.

For each event and for each feature, we update our EMA histogram using Algorithm \ref{pseudo:ema-hist} which has to perform a binary search to find the correct bin to insert our new value and traverses all bins to apply a suppression factor. Using $n$ as the number of bins, we have a time complexity of $O(log(n) + n)$, for each feature. However, since $n$ is a fixed and small constant, \textit{i.e.}, more or less 100 bins, we say that this algorithm has constant time complexity. In other words, our time complexity does not depend on the size of the emulated streaming windows.

\subsubsection*{Space complexity of the streaming phase}
We store $n$ bins per feature, so we have a space complexity of $O(n)$ per feature. Again, since $n$ is a small and fixed constant, we claim to have constant space complexity for ever-growing volumes of data.

>>>>

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
Many real-time stream monitoring systems are static once deployed in a production environment. Over time, due to data pattern shifts, the initially deployed static system's performance gradually deteriorates. Data pattern shift detection refers to the process of finding patterns in data that do not conform to expected or usual behavior. Accurate and timely detection of data pattern deviations allow for immediate measures to be taken. Thus, the problem at hand is to determine when to reconfigure the system, \textit{e.g.}, a Machine Learning model, based on an analysis of the drifts in the stream of data. In this paper, we present a method that makes use of constant time & memory streaming aggregations and distribution divergence functions to alert deviations in data patterns in real-time. We generate alerts based on a user-defined probability threshold, for each of an event's fields, denominated features in our context. We evaluate our method through a series of tests, some with synthetic data sets and some with real data. We further split the tests into single and multi-feature analysis. Our method accurately detected the introduced anomalies in the experiments using synthetic datasets while maintaining high throughput, for single and multi-feature analysis. However, experiments with real data were not as accurate. Despite not knowing the specific reasons for that, we detail further investigation to future work and formulate a set of hypotheses that might explain the latter results and hence are worthy of pursuing next. Moreover, our set of experiments support the claim that sliding window aggregations and distribution divergence methods can be combined to detect data pattern shifts in streaming scenarios with constant time and memory complexity.
\end{abstract}

@@

\section{Method}
\label{sec:Method}
A data stream is a continuous collection of events, each timestamped and with multiple fields. In this paper, we refer to the event's fields as features. Our goal is to detect data pattern deviations for each feature in a streaming fashion. To that end, we devise a method that works with two windows and comprises two phases, a batch analysis phase and a stream analysis one. The batch analysis is performed on a static in time window which we call the reference window. The stream analysis is performed over a sliding window that groups data stream events. We aggregate the contents of the reference and sliding windows using an approximate histogram aggregation based on Exponential Moving Averages (EMAs) \cite{EMA-Everett2011, EMA-Hunter, EMA-MarcusB}. We then apply a statistical distribution divergence function to both aggregations to obtain a divergence value. In the batch phase, we build the distribution of these values so we can define a percentile threshold for newly measured divergence values.

\subsection{Batch Analysis Phase} \label{sec:batch-phase}
The goal of the batch phase is to build our reference aggregation for each feature and get to know the distribution of divergence or distance values. In other words, for each feature, we:
\begin{enumerate}
    \item Compute the reference histogram using all data in the reference period
    \item Compute the distribution of distances between the reference histogram and random samples of data
\end{enumerate}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/ref-hist.png}
      \caption[]{Reference equal-height histogram}
      \label{fig:ref-hist}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/ref-bins.png}
      \caption[]{Reference equal-height histogram}
      \label{fig:ref-hist-bins}
    \end{center}
\end{figure}

Each of the histograms is an EMA-like histogram. An EMA-like histogram is essentially a collection of EMA-counts, as described in Section \ref{sec:ema-hist}. Each sample is processed event by event and a discount factor is applied per event to each bin (EMA expiration factor), thus building an approximated histogram aggregation, using the reference bins, for each feature of that sample. This procedure is illustrated in Figure \ref{fig:EMA-hist-build}.
\begin{figure}[!htb]
    \begin{center}
    \hspace{2cm}
      \includegraphics[scale=0.8]{figures/ema-hist.png}
      \caption[]{Reference equal-height histogram}
      \label{fig:EMA-hist-build}
    \end{center}
\end{figure}
We compute multiple sample histograms to encode distributions over smaller time periods that may exist within the large reference period. We use an approximated histogram to mimic the target approximated histogram we will incrementally maintain in the streaming environment.

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.7]{figures/compute-sample-distances.png}
      \caption[Compute sample's distance values]{Compute as many distance values as samples, one for each sample-reference histogram pair}
      \label{fig:compute-sample-distances}
    \end{center}
\end{figure}

@@

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/skewed-left-distro.png}
      \caption[]{Left skewed distribution of distance values}
      \label{fig:skewed-left-distro}
    \end{center}
\end{figure}

\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.4]{figures/skewed-right-distro.png}
      \caption[]{Right skewed distribution of distance values}
      \label{fig:skewed-right-distro}
    \end{center}
\end{figure}

@@
